SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-11-19 2017-Xudong-SMT/ActaInformatica.tex
Adjust the position of table on page 25.
a couple of typos fixed

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined,  e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, they will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y_1. a_1\neq \delta(y_1)]$. It denotes a possible behaviour of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is, any action that does not match $\delta(y_1)$, for any value of the variable $y_1$.
Remark that the data-oriented encoding in pNet $EnableData$ has the same synchronisation vectors, up to the usual alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

Figure~\ref{schema:ArchFailure:BIP} shows this modified version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.\footnote{%
%
  \Simon{%
  The original CubETH case study \cite{CubETH-case-study} focused on
  the possibility of assembling a model of a complex software system
  in a systematic way by applying architectures to discharge
  individual system requirements.  Here, we are more interested by the
  properties of the Failure Monitor architecture by itself.  For that
  reason, we provide a modified version, in particular decomposing the
  coordinator into a Controller and a Timer, altering them to allow
  more flexibility in the acceptable behaviours.  In other words, this
  modification enforces essentially the same properties, while
  discarding less acceptable behaviours.
  }
%
} The architecture consists of 2 \emph{coordinator components}
($\NameCtrl$ and $\NameTimer$, for \emph{Control} and \emph{Timer},
respectively), 5 \emph{dangling ports} ($\PortFail$, $\PortResume$,
$\PortTick$, $\PortReset$ and $\PortAsk$) and 5 connectors.
%
%% Contrary to standard BIP models, architectures comprise one or several
%% \emph{operand} components, whereof only the set of \emph{ports} is
%% given.  Here, the operand component is $\NameOprnd$ and its interface
%% consists of the ports $\PortResume$ and $\PortFail$.
%% The two \emph{coordinator} components\mdash $\NameCtrl$ and
%% $\NameTimer$\mdash are standard
The  \emph{behaviour} of BIP components %% insofar as they also
%% have their 
is specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  %% For instance, the loop transition
%% $t_1 \goesto{\PortTick, [t < Max], t:=t+1} t_1$
%% in the $\NameTimer$ component is labelled by the port $\PortTick$. It
%% can be fired only when the current value of the local variable $t$ is
%% lower than $Max$.  Upon firing, this transition increments the value
%% of $t$ by $1$.  When omitted, the default guard (\resp update
%% function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
%% operator).  The constant $\mathrm{Max}$ above, is a parameter of the
%% architecture.
Although this is not visible in the figure, below we will assume that
the dangling ports $\PortFail$ and $\PortResume$ belong to the operand
component, whereas the dangling ports $\PortTick$, $\PortAsk$ and
$\PortReset$ represent actions of the environment.

@@

The satisfaction of the safety property (\ref{eq:FM:safety}) could be
established by applying symbolic model checking techniques.  However,
in this example, it is obvious by inspection of the open automaton.
The satisfaction of the liveness property relies on the observation
that \Simon{the only loop starting in the state $\ostate{s1,t1}$ and involving neither $\mathit{reset}$, nor $\mathit{resume}$ or $\mathit{ask}$, is the self-loop $\mathit{fail(b0)}$.  By further inspecting the synchronisation vector $SV0$, we observe that this self-loop corresponds to the transitions $\mathit{fail(false)}$ and $\mathit{start(false)}$ in the two pLTSs.  Recall that these transitions (carrying the value $\mathit{false}$) are introduced by the encoding. That is, none of the two coordinating components are involved in the joint transition.  Therefore, under resonable scheduling assumptions,
the transition $\mathit{ask}$ will eventually be fired.}

@@

We validate the approach on \Simon{a larger example},%\st{two larger examples}, 
based on a BIP architecture taken from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  \Simon{This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.
In order to evaluate the potential
%
of our approach to scale to larger systems, we also apply the proposed
techniques to a BIP architecture obtained by combining two instances
of the Failure Monitor architecture.  Our results show that, even
though, the performance of the smart algorithm degrades more than
linearly, execution times remain reasonably small.  It should be noted
that the aim of our approach, particularly in the case of BIP
architectures, is to allow symbolic verification of design patterns to
show that they do impose corresponding properties. Safety of the system is obtained compositionally, whereas liveness is verified by checking \emph{pair-wise} non-interference. In both cases, the main characteristic of the approach is compsitionality, so we can compute the behavior and analyse the properties on \emph{small open systems}, and use these properties independently when composing bigger systems.
%
}
%% \st{We also show how such architectures
%% can be composed, and how our approach scales.
%% This example shows that
%% open-automata-based semantics can be instrumental in verifying the
%% properties enforced by the architectures through an encoding into open
%% pNets.
%% We also compare the practical performances of the two algorithms and
%% show the benefit gained using the smart approach.}

@@

\Simon{BIP architectures in this paper and in the CubETH case study \cite{CubETH-case-study} involved data in the BIP components behaviour but not in the interactions among components.  We have shown in \cite{BHM19:Coordination} how to overcome this limit, allowing data transfer in BIP architecture connectors and providing an encoding of such architectures in terms of open pNets.  To push the study of BIP architecture verification using open pNets even further, one can observe that there is a certain ``regularity'' in the structure of the composed architecture in Figure~\ref{schema:ArchFailure:BIP:Double}.  Indeed, both architectures in Figures~\ref{schema:ArchFailure:BIP} and \ref{schema:ArchFailure:BIP:Double} are instances of the same \emph{architecture style} \cite{MBBS16-diagrams} for 1 and 2 monitored processes, respectively.  The question then naturally arises of whether some version of open pNets (\eg extended to accommodate a parameterised number of sub-pNets) can be used to verify certain architectural styles without instantiating them with fixed parameter values.  Although this problem is know to be undecidable in the general case (see, \eg \cite{KKW+16-foil-concur}), it would be interesting to study the interplay of existing techniques, such as well structured transition systems \cite{WSTSE!}, with open pNet extensions.}

>>>>

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined,  e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, they will be crucial for defining weak equivalences. The pNet can be given as follows:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y_1. a_1\neq \delta(y_1)]$. It denotes a possible behaviour of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is, any action that does not match $\delta(y_1)$, for any value of the variable $y_1$.
Remark that the data-oriented encoding in pNet $EnableData$ has the same synchronisation vectors, up to the usual alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

Figure~\ref{schema:ArchFailure:BIP} shows this modified version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.\footnote{%
%
  \Simon{%
  The original CubETH case study \cite{CubETH-case-study} focused on
  the possibility of assembling a model of a complex software system
  in a systematic way by applying architectures to discharge
  individual system requirements.  Here, we are more interested in the
  properties of the Failure Monitor architecture by itself.  For that
  reason, we provide a modified version, in particular decomposing the
  coordinator into a Controller and a Timer, altering them to allow
  more flexibility in the acceptable behaviours.  In other words, this
  modification enforces essentially the same properties, while
  discarding less acceptable behaviours.
  }
%
} The architecture consists of 2 \emph{coordinator components}
($\NameCtrl$ and $\NameTimer$, for \emph{Control} and \emph{Timer},
respectively), 5 \emph{dangling ports} ($\PortFail$, $\PortResume$,
$\PortTick$, $\PortReset$ and $\PortAsk$) and 5 connectors.
%
%% Contrary to standard BIP models, architectures comprise one or several
%% \emph{operand} components, whereof only the set of \emph{ports} is
%% given.  Here, the operand component is $\NameOprnd$ and its interface
%% consists of the ports $\PortResume$ and $\PortFail$.
%% The two \emph{coordinator} components\mdash $\NameCtrl$ and
%% $\NameTimer$\mdash are standard
The  \emph{behaviour} of BIP components %% insofar as they also
%% have their 
is specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  %% For instance, the loop transition
%% $t_1 \goesto{\PortTick, [t < Max], t:=t+1} t_1$
%% in the $\NameTimer$ component is labelled by the port $\PortTick$. It
%% can be fired only when the current value of the local variable $t$ is
%% lower than $Max$.  Upon firing, this transition increments the value
%% of $t$ by $1$.  When omitted, the default guard (\resp update
%% function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
%% operator).  The constant $\mathrm{Max}$ above, is a parameter of the
%% architecture.
Although this is not visible in the figure, below we will assume that
the dangling ports $\PortFail$ and $\PortResume$ belong to the operand
component, whereas the dangling ports $\PortTick$, $\PortAsk$ and
$\PortReset$ represent actions of the environment.

@@

The satisfaction of the safety property (\ref{eq:FM:safety}) could be
established by applying symbolic model checking techniques.  However,
in this example, it is obvious by inspection of the open automaton.
The satisfaction of the liveness property relies on the observation
that \Simon{the only loop starting in the state $\ostate{s1,t1}$ and involving neither $\mathit{reset}$, nor $\mathit{resume}$ or $\mathit{ask}$, is the self-loop $\mathit{fail(b0)}$.  By further inspecting the synchronisation vector $SV0$, we observe that this self-loop corresponds to the transitions $\mathit{fail(false)}$ and $\mathit{start(false)}$ in the two PLTSs.  Recall that these transitions (carrying the value $\mathit{false}$) are introduced by the encoding. That is, none of the two coordinating components are involved in the joint transition.  Therefore, under resonable scheduling assumptions,
the transition $\mathit{ask}$ will eventually be fired.}

@@

We validate the approach on \Simon{a larger example}, %\st{two larger examples}, 
based on a BIP architecture taken from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  \Simon{This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.
In order to evaluate the potential
%
of our approach to scale to larger systems, we also apply the proposed
techniques to a BIP architecture obtained by combining two instances
of the Failure Monitor architecture.  Our results show that, even
though, the performance of the smart algorithm degrades more than
linearly, execution times remain reasonably small.  It should be noted
that the aim of our approach, particularly in the case of BIP
architectures, is to allow symbolic verification of design patterns to
show that they do impose corresponding properties. Safety of the system is obtained compositionally, whereas liveness is verified by checking \emph{pair-wise} non-interference. In both cases, the main characteristic of the approach is compsitionality, so we can compute the behavior and analyse the properties on \emph{small open systems}, and use these properties independently when composing bigger systems.
%
}
%% \st{We also show how such architectures
%% can be composed, and how our approach scales.
%% This example shows that
%% open-automata-based semantics can be instrumental in verifying the
%% properties enforced by the architectures through an encoding into open
%% pNets.
%% We also compare the practical performances of the two algorithms and
%% show the benefit gained using the smart approach.}

@@

\Simon{BIP architectures in this paper and in the CubETH case study \cite{CubETH-case-study} involved data in the BIP components behaviour but not in the interactions among components.  We have shown in \cite{BHM19:Coordination} how to overcome this limit, allowing data transfer in BIP architecture connectors and providing an encoding of such architectures in terms of open pNets.  To push the study of BIP architecture verification using open pNets even further, one can observe that there is a certain ``regularity'' in the structure of the composed architecture in Figure~\ref{schema:ArchFailure:BIP:Double}.  Indeed, both architectures in Figures~\ref{schema:ArchFailure:BIP} and \ref{schema:ArchFailure:BIP:Double} are instances of the same \emph{architecture style} \cite{MBBS16-diagrams} for 1 and 2 monitored processes, respectively.  The question then naturally arises of whether some version of open pNets (\eg extended to accommodate a parameterised number of sub-pNets) can be used to verify certain architectural styles without instantiating them with fixed parameter values.  Although this problem is known to be undecidable in the general case (see, \eg \cite{KKW+16-foil-concur}), it would be interesting to study the interplay of existing techniques, such as well structured transition systems \cite{WSTSE!}, with open pNet extensions.}
